Wireless Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data

### 2011-09-23 docs/thesis/sources/literature_survey.tex
Final thesis draft.

There are several routing protocols currently used in WSNs~\cite{survey:wirelessrouting}, unfortunately, most are not well suited for BSNs. Several of the protocols focus on mobility, Quality of Service (QoS), scalability and reliability, among other things. In general, BSNs do not have the same requirements as other WSNs. For example, in BSNs, scalability is no longer a problem since networks are limited to a human body. In most cases, power consumption becomes the main limiting factor for overall system performance.

Some routing protocols are specifically designed with BSN constraints in mind. One example is the Wireless Autonomous Spanning tree Protocol (WASP) \cite{protocol:WASP}. Braem et al. describe the WASP as ``a slotted protocol that uses a spanning tree for medium access coordination and traffic routing''. The main idea is that each node sends a ``WASP-scheme'' through a broadcast message that both parent and children nodes receive. This scheme contains the time allotments for each node for the current time frame. The children of the current node need the scheme to know when they are allowed to transmit data. The scheme also informs the parent of the current node as to how many children each node has, which allows it to allocate enough time slots for that branch. This method allows for a distributed time-allocation protocol. Each time frame also includes contention periods where new nodes can join the network. Routing is simple as data from sensor nodes is just sent up the network tree and broadcast messages are used to send configuration down from the root node. The main disadvantage of the WASP is the need to send the scheme packets from the root node down the tree during every time frame.

@@

Other BSN routing methods focus on postural mobility~\cite{routing:storeandforward}. Quwaider et al. propose a protocol tolerant to constant network changes due to body/sensor movement. When sensor nodes are located in body extremities, the wireless links vary with body position changes. The proposed protocol uses a store-and-forward method to transfer data from sensor nodes to the main node. While the protocol is well suited for dealing with mobile nodes, it requires more energy to do so. ``Hello'' packets are constantly sent by each node to determine what the current neighboring nodes are. A neighbor table is generated using this information and is necessary to determine where packets will be sent to. Another problem with the method is due to multiple copies of packets being transmitted to increase the probability of delivery. Finally, the main node transmits at full power in order to poll each sensor node, increasing its energy consumption.

Most current routing methods for BSNs use network trees. One of the main problems present with network trees is that the nodes closer to the main node deplete their energy supply faster than children nodes. This is due to the fact that all trafic from children nodes has to be relayed through. Ehyaie et al. propose using dedicated relay nodes instead of having sensor nodes act as relays~\cite{relay:networklife}. Using dedicated relays increases the individual sensor life, with the added cost of more devices.

So far, most BSNs have used tree-like networks to join nodes together. Tree networks, while simple, are negatively affected by node failures and movement.  Nabi et al. use a gossiping data routing strategy along with TDMA based MAC to transmit information to the main node \cite{relay:transmitpoweradaptation}. Each node receives and stores the last sample data from every neighboring node. During transmission, it sends its own sampled data, along with one or more of the stored packets from other nodes. This method maximizes the likelihood of a packed reaching the main node, while consuming more power due to retransmissions of the same data. Timestamps are used to decide whether or not a received packed will be stored or discarded. The main node is assumed to have higher battery capacity and transmission range. It is required to be able to reach all nodes with a single transmission. It is in charge of sending a synchronization beacon to use as a time reference for TDMA frames. This method is not always feasible, for example, when two nodes are on opposite sides of the body.

@@

%\subsection{Similar Work}
%In order to determine the improvements of novel methods, reference systems need to be available to compare with. Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying BSN. Both proposed systems present, in detail, a TDMA MAC protocol specifically designed with energy efficiency in mind. Details of packet format, TDMA frame calculation, and power consumption measurements are available.
%Add more details later

>>>>

There are several routing protocols currently used in WSNs~\cite{survey:wirelessrouting}, unfortunately, most are not well suited for WBANs. Several of the protocols focus on mobility, Quality of Service (QoS), scalability and reliability, among other things. In general, WBANs do not have the same requirements as other WSNs. For example, in WBANs, scalability is no longer a problem since networks are limited to a human body. In most cases, power consumption becomes the main limiting factor for overall system performance.

Some routing protocols are specifically designed with WBAN constraints in mind. One example is the Wireless Autonomous Spanning tree Protocol (WASP) \cite{protocol:WASP}. Braem et al. describe the WASP as ``a slotted protocol that uses a spanning tree for medium access coordination and traffic routing''. The main idea is that each node sends a ``WASP-scheme'' through a broadcast message that both parent and children nodes receive. This scheme contains the time allotments for each node for the current time frame. The children of the current node need the scheme to know when they are allowed to transmit data. The scheme also informs the parent of the current node as to how many children each node has, which allows it to allocate enough time slots for that branch. This method allows for a distributed time-allocation protocol. Each time frame also includes contention periods where new nodes can join the network. Routing is simple as data from sensor nodes is just sent up the network tree and broadcast messages are used to send configuration down from the root node. The main disadvantage of the WASP is the need to send the scheme packets from the root node down the tree during every time frame.

@@

Other WBAN routing methods focus on postural mobility~\cite{routing:storeandforward}. Quwaider et al. propose a protocol tolerant to constant network changes due to body/sensor movement. When sensor nodes are located in body extremities, the wireless links vary with body position changes. The proposed protocol uses a store-and-forward method to transfer data from sensor nodes to the main node. While the protocol is well suited for dealing with mobile nodes, it requires more energy to do so. ``Hello'' packets are constantly sent by each node to determine what the current neighboring nodes are. A neighbor table is generated using this information and is necessary to determine where packets will be sent to. Another problem with the method is due to multiple copies of packets being transmitted to increase the probability of delivery. Finally, the main node transmits at full power in order to poll each sensor node, increasing its energy consumption.

Most current routing methods for WBANs use network trees. One of the main problems present with network trees is that the nodes closer to the main node deplete their energy supply faster than children nodes. This is due to the fact that all trafic from children nodes has to be relayed through. Ehyaie et al. propose using dedicated relay nodes instead of having sensor nodes act as relays~\cite{relay:networklife}. Using dedicated relays increases the individual sensor life, with the added cost of more devices.

So far, most WBANs have used tree-like networks to join nodes together. Tree networks, while simple, are negatively affected by node failures and movement.  Nabi et al. use a gossiping data routing strategy along with TDMA based MAC to transmit information to the main node \cite{relay:transmitpoweradaptation}. Each node receives and stores the last sample data from every neighboring node. During transmission, it sends its own sampled data, along with one or more of the stored packets from other nodes. This method maximizes the likelihood of a packed reaching the main node, while consuming more power due to retransmissions of the same data. Timestamps are used to decide whether or not a received packed will be stored or discarded. The main node is assumed to have higher battery capacity and transmission range. It is required to be able to reach all nodes with a single transmission. It is in charge of sending a synchronization beacon to use as a time reference for TDMA frames. This method is not always feasible, for example, when two nodes are on opposite sides of the body.

@@

%\subsection{Similar Work}
%In order to determine the improvements of novel methods, reference systems need to be available to compare with. Both~\cite{mac:tdma}~and~\cite{mac:lowdutycycle} provide a suitable reference system for a synchronized, non-relaying WBAN. Both proposed systems present, in detail, a TDMA MAC protocol specifically designed with energy efficiency in mind. Details of packet format, TDMA frame calculation, and power consumption measurements are available.
%Add more details later

### 2011-09-23 docs/thesis/thesis.tex
Final thesis draft.

\begin{document}
\title{Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data}
\author{Alvaro Prieto\\
Masters Thesis\\
BS/MS Electrical Engineering\\
Electrical and Microelectronic Engineering Department\\
Rochester Institute of Technology\\
Rochester, NY\\
\texttt{axp8664@rit.edu}}

@@

\subsection{Design Choices}
The decision to write all software libraries from scratch was not an easy one. Initially the SimpliciTI libraries were modified to try and meet the project needs, but this proved to be a complex task. Due to the simple nature of the planned WBAN, a much less complex approach could be used. The main application only required streaming data from various wireless sensors.$\ll$ Should I elaborate on the design constraints here? The number of devices is known, the time of transmission is known etc...$\gg$ This could be accomplished with a very minimal radio library.

@@

The host is a laptop computer with an Intel(R) Core(TM) i5-2410M CPU and 4GB of RAM. Eight EDs were placed on a 170cm, 70kg male subject as shown in Figure~\ref{fig:body}. The host was carried on a backpack and connected to the AP via a USB cable. The data was sampled, and the routing algorithm was run, at a rate of 5Hz. This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.

@@

\begin{table}[htb]
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
\hline 
\end{tabular} 
\caption{Sample RSSI Table}
\label{tab:rssi}
\end{table}

@@

\begin{figure}[!htb]
\includegraphics[width=0.8\textwidth]{figures/sit1-c0.pdf}
\caption{Sitting M=0}
\label{fig:sit1-c0}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.8\textwidth]{figures/sit1-c100.pdf}
\caption{Sitting M=100}
\label{fig:sit1-c100}
\end{figure}

@@

Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime. For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.7 mW. On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 1.3 mW each. One important observation is that the network with $M=0$ consumed less overall power (~9.6mW) than the one with $M=100$ (~10.1mW). This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network. While more energy is consumed overall, no single ED consumes much more power than others.

@@

To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 1mW of normalized energy. In the previous example, the trial with $M=0$ reaches 1mW after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.

\begin{figure}[!htb]
\includegraphics[width=0.8\textwidth]{figures/walk2-c0.pdf}
\caption{Walking M=0}
\label{fig:walk2-c0}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=0.8\textwidth]{figures/walk2-c100.pdf}
\caption{Walking M=100}
\label{fig:walk2-c100}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=0.8\textwidth]{figures/walk4-c100.pdf}
\caption{Walking M=100}
\label{fig:walk4-c100}
\end{figure}

>>>>

\begin{document}
\title{Wireless Body Area Network Platform Utilizing Energy-Efficient Routing of Physiological Data}
\author{by\\
Alvaro Prieto\\
Master of Science Thesis\\
in\\
Electrical Engineering\\
Approved by:\\
\makebox[3.5in]{\hrulefill}\\
(Gill Tsouri, Ph. D.)\\
\makebox[3.5in]{\hrulefill}\\
(Andres Kwasinski, Ph. D.)\\
\makebox[3.5in]{\hrulefill}\\
(Nirmala Shenoy, Ph. D.)\\
\makebox[3.5in]{\hrulefill}\\
(Sohail A. Dianat, Ph. D.)\\
Electrical and Microelectronic Engineering Department\\
Kate Gleason College of Engineering\\
Rochester Institute of Technology\\
Rochester, New York\\
\texttt{axp8664@rit.edu}}

@@

\subsection{Design Choices}
The decision to write all software libraries from scratch was not an easy one. Initially the SimpliciTI libraries were modified to try and meet the project needs, but this proved to be a complex task. Due to the simple nature of the planned WBAN, a much less complex approach could be used. The main application only required streaming data from various wireless sensors. This could be accomplished with a very minimal radio library.

@@

The host is a laptop computer with an Intel(R) Core(TM) i5-2410M CPU and 4GB of RAM. Eight EDs were placed on a 170cm, 70kg male subject as shown in Figure~\ref{fig:body}. The host was carried on a backpack and connected to the AP via a USB cable. The data was sampled, and the routing algorithm was run, at a rate of 5Hz. This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this. If data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.

@@

\begin{table}[htb]
\begin{center}

\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
\hline 
\end{tabular} 
\caption{Sample RSSI Table}
\label{tab:rssi}
\end{center}
\end{table}

@@

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/sit1-c0.pdf}
\caption{Sitting M=0}
\label{fig:sit1-c0}
\end{center}
\end{figure}

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/sit1-c100.pdf}
\caption{Sitting M=100}
\label{fig:sit1-c100}
\end{center}
\end{figure}

@@

Figures~\ref{fig:walk2-c0}~and~\ref{fig:walk2-c100}, on the other hand, show an increase in network lifetime. For example, at the end of the 5 minute sampling run with $M=0$, the device which consumed the most power was ED7 with approximately 1.06 $\mu$J. On the other hand, with $M=100$, all EDs consume power in a uniform fashion and end up with less than 0.81 $\mu$J each. One important observation is that the network with $M=0$ consumed less overall power (~6 $\mu$J) than the one with $M=100$ (~6.31$\mu$J). This is expected, due to the fact that the new cost function specifically avoids the most efficient path for a specific node in favor of the most efficient path for the whole network. While more energy is consumed overall, no single ED consumes much more power than others.

@@

To demonstrate the network lifetime improvement, an ED will be considered have depleted it's power source after consuming 0.625 $\mu$J of energy. In the previous example, the trial with $M=0$ reaches 0.625 $\mu$J after approximately 175 seconds, while the case with $M=100$ reaches it around 240 seconds.

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/walk2-c0.pdf}
\caption{Walking M=0}
\label{fig:walk2-c0}
\end{center}
\end{figure}

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/walk2-c100.pdf}
\caption{Walking M=100}
\label{fig:walk2-c100}
\end{center}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/walk4-c100.pdf}
\caption{Walking M=100}
\label{fig:walk4-c100}
\end{center}
\end{figure}
